Clinical advancement forecasting \thanks{\textit{\underline{Citation}}: \textbf{TBD}}

### 2024-06-06 paper/paper.tex
More review revisions

%% Authors
\author{
  Eric Czech\thanks{Correspondence to Eric Czech (eric@related.vc).} \hspace{0.2cm} 
  Rafal Wojdyla \hspace{0.2cm} 
  Daniel Himmelstein \hspace{0.2cm} 
  \\ \\
  \textbf{Jack Milwid} \hspace{0.2cm} 
  \textbf{Adam Kolom} \hspace{0.2cm} 
  \textbf{Jeff Hammerbacher}
  \\ \\
  Related Sciences LLC
}

@@

It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{PMID:24833294,PMID:35804044,Razuvayevskaya2023.02.07.23285407,PMID:37803084,PMID:36963162}, typically by a factor of at least 2x \cite{Minikel2023.06.23.23291765,Nelson2015-eg,King2019-rc,PMID:30652614,Sadler2023-xd}. This information has been used to devise target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}. It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}. \ul{To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates}. In contrast, we aim to establish a framework for identifying targets-disease (TD) pairs with the very highest possible likelihood of success in order to establish an evidence-based method to optimize research and development pipelines. This is accomplished in this study by integrating a subset of our evidence base from Open Targets \cite{Koscielny2017-rr} spanning human clinical, genomic, transcriptomic and proteomic data as well as cell and animal model evidence, pathway information, and basic literature metrics, in a simple statistical modeling framework.

@@

Precedent studies and methods discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with observed clinical trial success, and 2) clinical trial outcome prediction models. Both are measured against the same outcomes and an important distinction between them lies in how prioritization methods are \textbf{not} directly optimized for those outcomes while outcome prediction methods are. \ul{In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial}. We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS. Finally, we examine how many present-day target-disease pairs are likely to see success rates matching or exceeding those calibrated benchmarks, that are also undeveloped (i.e. have never been in clinical trials) and have a tractable target.

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \captionsetup[subfigure]{labelformat=empty}
  \subfloat[\centering]{
    \begin{overpic}[width=.9\textwidth]{feature_presence_unscaled.pdf}
      \put(102, 20) {(a)}
    \end{overpic}
  }
  \qquad
  \subfloat[\centering]{
    \begin{overpic}[width=.9\textwidth]{feature_presence_scaled.pdf}
      \put(102, 20) {(b)}
    \end{overpic}
  }
  \caption{
    \textbf{Evaluation dataset feature presence}.
    (a) Presence and co-occurrence of select features for overlapping sets of TD pairs. The size of each bar is proportional to the number of TD pairs with a given combination of features and the sort ordering is determined by the number of TD pairs associated with any one feature (show in the right margin). The top margin shows the number of features in any one feature combination and the bottom margin shows the number of TD pairs that are a part of that same combination.
    (b) Same as (a) with proportional bar scaling that does not drop beyond a limit necessary to fit all margin labels and combinatorial groupings. Features that are not specific to TD pairs are omitted for brevity, along with the outcome feature used in this study and denoted here with the label "pair has advanced beyond phase 2".
  }
  \label{fig:feature_presence}
\end{figure}

@@

We develop our predictive framework using two primary types of models: linear models, where relationships between features and outcomes are assumed to be linear, and gradient-boosted tree models, which can capture non-linear relationships. Both models were tested in constrained formats, ensuring all feature effects are non-negative, and unconstrained formats, allowing for negative relationships. This is possible with no underlying feature transformations because all scores in Open Targets are constructed such that higher scores are presumed to be advantageous. We also note that neural models are a significant omission from this framework. While deep learning methods for tabular data are competitive with gradient-boosted tree models \cite{mcelfresh2023neural, gorishniy2023revisiting, somepalli2021saint}, they are rarely superior without transfer learning \cite{mcelfresh2023neural, levin2023transfer, hollmann2023tabpfn}. They also have a number of operational drawbacks like necessitating GPUs and having less standard hyperparameter tuning procedures. Additionally, the data in this proof-of-concept is limited to Open Targets evidence scores that already reflect a large degree of feature engineering, thereby limiting the extent to which deep learning networks can identify more granular predictive signals.

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \captionsetup[subfigure]{labelformat=empty}
  \subfloat[\centering]{
    \begin{overpic}[width=.57\textwidth]{relative_risk_mean_across_ta.pdf}
      \put(50, -3) {(a)}
      % Performance multipliers
      \put(10.5,28.5){\line(0,1){27.5}} % Vertical line
      \put(10.5,28.5){\line(1,0){.5}} % Bottom bar
      \put(10.5,56){\line(1,0){.5}} % Top bar
      \put(6,42.5){\sffamily \fontsize{8pt}{12pt}\selectfont 2x}
      \put(30.5,22.25){\line(0,1){6.5}}
      \put(23,25){\sffamily \fontsize{8pt}{12pt}\selectfont 1.5x}
      % Ranking percentages
      \put(19,9){\line(0,1){4}}
      \put(15,14){\sffamily \fontsize{8pt}{12pt}\selectfont Top 1\%}
      \put(36.5,9){\line(0,1){4}}
      \put(32.5,14){\sffamily \fontsize{8pt}{12pt}\selectfont Top 2\%}
    \end{overpic}
  }
  \qquad
  \subfloat[\centering]{
    \begin{overpic}[width=.36\textwidth]{classifier_metrics_dist_across_ta.pdf}
      \put(27, -3) {(b)}
    \end{overpic}
  }
  \caption{
    \textbf{Performance compared to Open Targets composite scores}.
    (a) Equally-weighted average relative risk estimates across 13 therapeutic areas, by number of top rankings and 3 methods: RDG (ours), RDG-T (ours) and OTS (Open Targets composite scores). 
    (b) Receiver operating characteristic (ROC) and average precision scores across the same 13 therapeutic areas with no limit on the number of rankings. 
    See Supplementary~Figure~\ref{fig:relative_risk_by_ta} for raw data underlying (a).
    Annotations indicate RR multiples between RDG and OTS for the top 10 and top 30 TD pairs as well as how many TD pairs comprise the top 1\% (N=18) and top 2\% (N=36) of rankings on average across therapeutic areas. All results shown are from the evaluation dataset.
    % Ranking percentages are based on an average of 1796 TD pairs per therapeutic area
  }
  \label{fig:performance_across_ta}
\end{figure}

@@

One key objective of this study is to determine if any model, e.g. RDG, can sort TD pairs with genetic support such that at least some portion of that sorted list has a likelihood of advancement that consistently exceeds what is expected from any one source of genetic support. We find that this goal is met and surpassed by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support and have an equivalent or greater expected rate of advancement. See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model. This latter, supplementary figure further emphasizes our focus on prioritizing opportunities beyond those with genetic support and provides examples of TD pairs with multiple lines of evidence.

@@

\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The RDG-T model achieves far higher ROC and AP scores through the use of the time since transition feature, which indicates the number of years a TD pair has been underdevelopment after having reached phase 2.
  \item The RDG model, however, matches or exceeds RDG-T in performance among top TD pairs
  \item The RDG-X model, using no human clinical or genetic evidence linked to a disease, outperforms the Open Targets composite score and nearly matches the performance of the RDG model beyond top rankings
  \item Linear models outperform gradient-boosting models by nearly all measures
  \item Constrained linear models outperform unconstrained linear models by nearly all measures
\end{enumerate}

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{performance_metric_mean_across_ta.pdf}
    \put(84.7, 38.2) {(a)}
    \put(84.7, 22.5) {(b)}
  \end{overpic}
  \caption{
    \textbf{Performance across model algorithms and feature ablation groups}.
    Average precision (AP) and receiver operating characteristic (ROC) scores with relative risk (RR) at ranking cutoffs denoted by \colorbox{Gainsboro}{RR@N}. 
    (a) Performance across constrained RDG models using different groups of features as described in Section~\ref{sec:models}
    (b) Performance across constrained (\colorbox{Gainsboro}{[+]}) and unconstrained (\colorbox{Gainsboro}{[+/-]}) linear and gradient-boosted models using the core feature set.
  }
  \label{fig:performance_metric_mean_across_ta}
\end{figure}

@@

\subsection{Opportunities}
\label{sec:results_opportunities}

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{opportunity_funnel.pdf}
    \put(100, 53) {(a)}
    \put(100, 41) {(b)}
    \put(100, 19) {(c)}
    \put(50, 51){\color{black}\vector(0, -1){5}}
    \put(74, 39){\color{black}\vector(0, -1){4.5}}
    \put(13.5, 58){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.53M TD pairs with evidence linking them, 2.46M have never been in clinical trials.}
    \put(13.5, 35.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.46M undeveloped pairs, the top 0.8\% are 2.07x+ more likely to advance.}
    \put(13.5, 1.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 8,560 pairs 2.93x+ more likely to advance, 2,408 have clinical small molecule targets (e.g.).}
  \end{overpic}
  \caption{
    \textbf{Present-day target-disease pair opportunity funnel}.
    (a) All 2.53M TD pairs in Open Targets with evidence of some kind grouped by the maximum clinical trial phase reached for each as of the present day. The \colorbox{Gainsboro}{Undeveloped} pairs have never been tested in clinical trials. TD pair counts are shown in parentheses. Labels for phase 1, 3 and 4 trials are omitted for brevity.
    (b) All 2.46M undeveloped TD pairs for which RDG model predictions exceed the thresholds determined by the corresponding genetic benchmarks in Figure~\ref{fig:relative_risk_by_limit}. The \colorbox{Gainsboro}{Medium} threshold was calibrated by genetic support across OTG (i.e. GWAS evidence). The associated RR of 2.07 implies that pairs with RDG predictions above this threshold are at least 2.07 more likely to advance clinically than those below the threshold. The same is true for the \colorbox{Gainsboro}{High} and \colorbox{Gainsboro}{Highest} thresholds, which correspond to thresholds determined by EVA and OMIM, respectively. These groupings are not mutually exclusive, e.g. the 20,821 TD pairs above the \colorbox{Gainsboro}{Medium} threshold are a superset of the pairs above the \colorbox{Gainsboro}{High} threshold.
    (c) All 8,560 undeveloped TD pairs with an expected clinical advancement rate of at least 2.93x grouped by tractability modality and confidence level. The \colorbox{Gainsboro}{Other} modality corresponds to any modality not shown that is associated with a target tested in clinical trials previously. The \colorbox{Gainsboro}{High} confidence grouping for this category means those targets have approved drugs. For each other modality, the \colorbox{Gainsboro}{High} confidence level corresponds to \textbf{any} level of clinical validation for drugs of that modality, not just approval. See Supplementary~Table~\ref{tab:tractability_buckets} for the definition of all modality confidence levels.
  }
  \label{fig:opportunity_funnel}
\end{figure}

@@

Lastly, we note that these TD pairs are identified using predictors that are not constrained by temporalization. In other words, the predictions used to identify \textbf{present-day} TD pairs with a high likelihood of clinical success are based on more evidence than what is available in other sections of this study. This is primarily because some evidence in Open Targets is not associated with publications. Avoiding that limitation is appropriate in this context.

@@

To address potential biases and inflation in our dataset, our methodology specifically considers the temporal emergence of evidence. By doing so, we ensure our predictions are grounded in data reflective of genuine biomedical insights, rather than artefacts of research focus or historical data accumulation trends.

@@

An important question, that remains difficult to answer, is how much target prioritization methods like the one explored in this study can improve the efficiency of drug discovery. One 2018 estimate suggests that improving success rates from phase 1 to approval by 1.4x (from 9.6\% to 13.8\%) would reduce the median R\&D cost for a \textbf{single drug} by \$480 million USD \cite{PMID:32125404}. If a 2x increase in success rates from phase 1 to approval could be expected when pursuing only genetically validated TD pairs \cite{Nelson2015-eg}, then this would increase to a \$686 million cost savings per drug. Pushing that even further to a 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively. These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_opportunities} that there are more than 8,000 such opportunities (i.e. untested, present-day TD pairs) where at least a 3x improvement in phase 2 success rates could be expected.

@@

\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item Open Targets does not currently provide drug approvals with associated dates. This is a part of the reason why we have chosen to focus only on advancement beyond phase 2, and not phase 3 as well. 
  \item Not all Open Targets evidence can be associated with publication dates. This evidence is ignored when used to construct longitudinal predictors; however, this limitation does not apply to Section~\ref{sec:results_opportunities}.
  \item The Open Targets composite score we use (OTS) is not equivalent to the overall composite score for TD pairs provided by Open Targets \cite{OTscoring}. The score they provide is not longitudinal. We create an approximation of it over time based on the same weighting scheme across evidence types, but without a harmonic sum aggregation function. Maximum scores for the same target, disease, year and evidence type combination are used instead.
  \item Estimating an absolute number of TD pairs associated with any criteria is difficult. In Section~\ref{sec:results_opportunities}, we present absolute counts of undeveloped TD pairs with tractable targets and a high likelihood of clinical success. These counts may be inflated by the fact that the associated targets and diseases are so sufficiently similar that they ought not to be counted as distinct examples. Defining a threshold for similarity to combat this problem is difficult and application-specific though, so we do not offer any solutions here.
  \item The source of OMIM associations we use is derived from a subset of associations available in EVA. This subset is defined by associations with a linked publication. This is only an approximation, but our internal research shows that 98\% of EVA records with linked publications also have at least one OMIM submission.
\end{enumerate}

>>>>

%% Authors
\author{
  Eric Czech\thanks{Correspondence to Eric Czech (eric@related.vc).} \hspace{0.2cm} 
  Rafal Wojdyla \hspace{0.2cm} 
  Daniel Himmelstein \hspace{0.2cm}
  Daniel Frank \hspace{0.2cm} 
  \\ \\
  \textbf{Jack Milwid} \hspace{0.2cm} 
  \textbf{Adam Kolom} \hspace{0.2cm} 
  \textbf{Jeff Hammerbacher}
  \\ \\
  Related Sciences LLC
}

@@

It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{PMID:24833294,PMID:35804044,Razuvayevskaya2023.02.07.23285407,PMID:37803084,PMID:36963162}, typically by a factor of at least 2x \cite{Minikel2023.06.23.23291765,Nelson2015-eg,King2019-rc,PMID:30652614,Sadler2023-xd}. This information has been used to devise target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}. It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}. \ul{To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates}. In contrast, we aim to establish a framework for identifying target-disease (TD) pairs with the very highest possible likelihood of success in order to establish an evidence-based method to optimize research and development pipelines. This is accomplished in this study by integrating a subset of our evidence base from Open Targets \cite{Koscielny2017-rr} spanning human clinical, genomic, transcriptomic and proteomic data as well as cell and animal model evidence, pathway information, and basic literature metrics, in a simple statistical modeling framework.

@@

Precedent studies and methods discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with observed clinical trial success, and 2) clinical trial outcome prediction models. Both are measured against the same outcomes and an important distinction between them lies in how prioritization methods are \textbf{not} directly optimized for those outcomes while outcome prediction methods are. \ul{In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial}. We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS. Finally, we examine how many present-day target-disease pairs are likely to see success rates matching or exceeding those calibrated benchmarks, that are also undeveloped (i.e. have never been in clinical trials) and have a tractable target (i.e. with evidence suggesting that therapeutic modulation is possible).

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \captionsetup[subfigure]{labelformat=empty}
  \subfloat[\centering]{
    \begin{overpic}[width=.9\textwidth]{feature_presence_unscaled.pdf}
      \put(102, 20) {(a)}
    \end{overpic}
  }
  \qquad
  \subfloat[\centering]{
    \begin{overpic}[width=.9\textwidth]{feature_presence_scaled.pdf}
      \put(102, 20) {(b)}
    \end{overpic}
  }
  \caption{
    \textbf{Evaluation dataset feature presence}.
    (a) Presence and co-occurrence of select features for overlapping sets of TD pairs. The size of each bar is proportional to the number of TD pairs with a given combination of features and the sort ordering is determined by the number of TD pairs associated with any one feature (show in the right margin). The top margin shows the number of features in any one feature combination and the bottom margin shows the number of TD pairs that are a part of that same combination. There are \evaluationDatasetPairCount\ TD pairs in the evaluation dataset overall.
    (b) Same as (a) with proportional bar scaling that does not drop beyond a limit necessary to fit all margin labels and combinatorial groupings. Features that are not specific to TD pairs are omitted for brevity, along with the outcome feature used in this study and denoted here with the label "pair has advanced beyond phase 2".
  }
  \label{fig:feature_presence}
\end{figure}

@@

We develop our predictive framework using two primary types of models: linear models, where relationships between features and outcomes are assumed to be linear, and gradient-boosted tree models, which can capture non-linear relationships. Both models were tested in constrained formats, ensuring all feature effects are non-negative, and unconstrained formats, allowing for negative relationships. This is possible with no underlying feature transformations because all scores in Open Targets are constructed such that higher scores are presumed to be advantageous. We also note that neural models are a significant omission from this framework. While deep learning methods for tabular data are competitive with gradient-boosted tree models \cite{mcelfresh2023neural, gorishniy2023revisiting, somepalli2021saint}, they are rarely superior without transfer learning \cite{mcelfresh2023neural, levin2023transfer, hollmann2023tabpfn}. Additionally, the data in this proof-of-concept is limited to Open Targets evidence scores that already reflect a large degree of feature engineering, thereby limiting the extent to which deep learning networks can identify more granular predictive signals.

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \captionsetup[subfigure]{labelformat=empty}
  \subfloat[\centering]{
    \begin{overpic}[width=.57\textwidth]{relative_risk_mean_across_ta.pdf}
      \put(50, -3) {(a)}
      % Performance multipliers
      \put(10.5,28.5){\line(0,1){27.5}} % Vertical line
      \put(10.5,28.5){\line(1,0){.5}} % Bottom bar
      \put(10.5,56){\line(1,0){.5}} % Top bar
      \put(6,42.5){\sffamily \fontsize{8pt}{12pt}\selectfont 2x}
      \put(30.5,22.25){\line(0,1){6.5}}
      \put(23,25){\sffamily \fontsize{8pt}{12pt}\selectfont 1.5x}
      % Ranking percentages
      \put(19,9){\line(0,1){4}}
      \put(15,14){\sffamily \fontsize{8pt}{12pt}\selectfont Top 1\%}
      \put(36.5,9){\line(0,1){4}}
      \put(32.5,14){\sffamily \fontsize{8pt}{12pt}\selectfont Top 2\%}
    \end{overpic}
  }
  \qquad
  \subfloat[\centering]{
    \begin{overpic}[width=.36\textwidth]{classifier_metrics_dist_across_ta.pdf}
      \put(27, -3) {(b)}
    \end{overpic}
  }
  \caption{
    \textbf{Performance compared to Open Targets composite scores}.
    (a) Equally-weighted average relative risk estimates across 13 therapeutic areas, by number of top rankings and 3 methods: RDG (ours), RDG-T (ours) and OTS (Open Targets composite scores). 
    (b) Receiver operating characteristic (ROC) and average precision scores across the same 13 therapeutic areas (each dot is one therapeutic area) with no limit on the number of rankings.
    See Supplementary~Figure~\ref{fig:relative_risk_by_ta} for raw data underlying (a).
    Annotations indicate RR multiples between RDG and OTS for the top 10 and top 30 TD pairs as well as how many TD pairs comprise the top 1\% (N=18) and top 2\% (N=36) of rankings on average across therapeutic areas. All results shown are from the evaluation dataset.
    % Ranking percentages are based on an average of 1796 TD pairs per therapeutic area
  }
  \label{fig:performance_across_ta}
\end{figure}

@@

One key objective of this study is to determine if any model, e.g. RDG, can sort TD pairs with genetic support such that at least some portion of that sorted list has a likelihood of advancement that consistently exceeds what is expected from any one source of genetic support. We find that this goal is met and surpassed by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support and have an equivalent or greater expected rate of advancement. See Section~\ref{sec:results_rankings} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model. This latter, supplementary figure further emphasizes our focus on prioritizing opportunities beyond those with genetic support and provides examples of TD pairs with multiple lines of evidence.

@@

\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The RDG-T model achieves far higher ROC and AP scores through the use of the time since transition feature, which indicates the number of years a TD pair has been under development after having reached phase 2.
  \item The RDG model, however, matches or exceeds RDG-T in performance among top TD pairs
  \item The RDG-X model, using no human clinical or genetic evidence linked to a disease, outperforms the Open Targets composite score and nearly matches the performance of the RDG model beyond top rankings
  \item Linear models outperform gradient-boosting models by nearly all measures
  \item Constrained linear models outperform unconstrained linear models by nearly all measures
\end{enumerate}

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{performance_metric_mean_across_ta.pdf}
    \put(84.7, 38.2) {(a)}
    \put(84.7, 22.5) {(b)}
  \end{overpic}
  \caption{
    \textbf{Performance across model algorithms and feature ablation groups}.
    Average precision (AP) and receiver operating characteristic (ROC) scores with relative risk (RR) at ranking cutoffs denoted by \colorbox{Gainsboro}{RR@N}. Dot sizes and colors (red=low score, green=high score) are both set according to the score shown in each cell.
    (a) Performance across constrained RDG models using different groups of features as described in Section~\ref{sec:models}
    (b) Performance across constrained (\colorbox{Gainsboro}{[+]}) and unconstrained (\colorbox{Gainsboro}{[+/-]}) linear and gradient-boosted models using the core feature set.
  }
  \label{fig:performance_metric_mean_across_ta}
\end{figure}

@@

\subsection{Rankings}
\label{sec:results_rankings}

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{opportunity_funnel.pdf}
    \put(100, 53) {(a)}
    \put(100, 41) {(b)}
    \put(100, 19) {(c)}
    \put(50, 51){\color{black}\vector(0, -1){5}}
    \put(74, 39){\color{black}\vector(0, -1){4.5}}
    \put(13.5, 58){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.53M TD pairs with evidence linking them, 2.46M have never been in clinical trials.}
    \put(13.5, 35.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.46M undeveloped pairs, the top 0.8\% are 2.07x+ more likely to advance.}
    \put(13.5, 1.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 8,560 pairs 2.93x+ more likely to advance, 2,408 have clinical small molecule targets (e.g.).}
  \end{overpic}
  \caption{
    \textbf{Present-day target-disease pair ranking funnel}.
    (a) All 2.53M TD pairs in Open Targets with evidence of some kind grouped by the maximum clinical trial phase reached for each as of the present day. The \colorbox{Gainsboro}{Undeveloped} pairs have never been tested in clinical trials. TD pair counts are shown in parentheses. Labels for phase 1, 3 and 4 trials are omitted for brevity.
    (b) All 2.46M undeveloped TD pairs for which RDG model predictions exceed the thresholds determined by the corresponding genetic benchmarks in Figure~\ref{fig:relative_risk_by_limit}. The \colorbox{Gainsboro}{Medium} threshold was calibrated by genetic support across OTG (i.e. GWAS evidence). The associated RR of 2.07 implies that pairs with RDG predictions above this threshold are at least 2.07 more likely to advance clinically than those below the threshold. The same is true for the \colorbox{Gainsboro}{High} and \colorbox{Gainsboro}{Highest} thresholds, which correspond to thresholds determined by EVA and OMIM, respectively. These groupings are not mutually exclusive, e.g. the 20,821 TD pairs above the \colorbox{Gainsboro}{Medium} threshold are a superset of the pairs above the \colorbox{Gainsboro}{High} threshold.
    (c) All 8,560 undeveloped TD pairs with an expected clinical advancement rate of at least 2.93x grouped by tractability modality and confidence level. The \colorbox{Gainsboro}{Other} modality corresponds to any modality not shown that is associated with a target tested in clinical trials previously. The \colorbox{Gainsboro}{High} confidence grouping for this category means those targets have approved drugs. For each other modality, the \colorbox{Gainsboro}{High} confidence level corresponds to \textbf{any} level of clinical validation for drugs of that modality, not just approval. See Supplementary~Table~\ref{tab:tractability_buckets} for the definition of all modality confidence levels.
  }
  \label{fig:opportunity_funnel}
\end{figure}

@@

Lastly, we note that these TD pairs are identified using predictors that are not constrained by temporalization. In other words, the predictions used to identify \textbf{present-day} TD pairs with a high likelihood of clinical success are based on more evidence than what is available in other sections of this study. This is primarily because some evidence in Open Targets is not associated with publications, which prohibits us from associating it with a timestamp. Avoiding that limitation is appropriate in this context.

@@

To address potential biases and inflation in our dataset, our methodology specifically considers the temporal emergence of evidence. By doing so, we ensure our predictions are grounded in data reflective of genuine biomedical insights, rather than artifacts of research focus or historical data accumulation trends.

@@

An important question, that remains difficult to answer, is how much target prioritization methods like the one explored in this study can improve the efficiency of drug discovery. One 2018 estimate suggests that improving success rates from phase 1 to approval by 1.4x (from 9.6\% to 13.8\%) would reduce the median R\&D cost for a \textbf{single drug} by \$480 million USD \cite{PMID:32125404}. If a 2x increase in success rates from phase 1 to approval could be expected when pursuing only genetically validated TD pairs \cite{Nelson2015-eg}, then this would increase to a \$686 million cost savings per drug. Pushing that even further to a 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively. These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_rankings} that there are more than 8,000 such opportunities (i.e. untested, present-day TD pairs) where at least a 3x improvement in phase 2 success rates could be expected.

@@

\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item Open Targets does not currently provide drug approvals with associated dates. This is a part of the reason why we have chosen to focus only on advancement beyond phase 2, and not phase 3 as well. 
  \item Not all Open Targets evidence can be associated with publication dates. This evidence is ignored when used to construct longitudinal predictors; however, this limitation does not apply to Section~\ref{sec:results_rankings}.
  \item The rankings of undeveloped TD pairs in Section~\ref{sec:results_rankings} are presented directly from Open Targets with no additional curation or vetting. We recognize that some of these TD pairs are for uninformative disease terms, e.g. the association between "CTLA4" and "autoimmune disease" in Supplementary~Figure~\ref{fig:top_opportunity_predictions}, and we opt not address these cases to best reflect the state of rankings within Open Targets. Several more computational steps are necessary in practice to identify a more enriched set of actionable TD pairs.
  \item The Open Targets composite score we use (OTS) is not equivalent to the overall composite score for TD pairs provided by Open Targets \cite{OTscoring}. The score they provide is not longitudinal. We create an approximation of it over time based on the same weighting scheme across evidence types, but without a harmonic sum aggregation function. Maximum scores for the same target, disease, year and evidence type combination are used instead.
  \item Estimating an absolute number of TD pairs associated with any criteria is difficult. In Section~\ref{sec:results_rankings}, we present absolute counts of undeveloped TD pairs with tractable targets and a high likelihood of clinical success. These counts may be inflated by the fact that the associated targets and diseases are so sufficiently similar that they ought not to be counted as distinct examples. Defining a threshold for similarity to combat this problem is difficult and application-specific though, so we do not offer any solutions here.
  \item The source of OMIM associations we use is derived from a subset of associations available in EVA. This subset is defined by associations with a linked publication. This is only an approximation, but our internal research shows that 98\% of EVA records with linked publications also have at least one OMIM submission.
\end{enumerate}
